---
title: "Printemps_tot"
author: "Aurel"
date: "2023-06-08"
output: html_document
---

# we load the data
```{r}
local_path <- "C:/Users/ahebertb/Desktop/StageAurel/"
source(paste(local_path,"code/script/boot.R",sep=""))

data_model <- readRDS(paste(local_path,"code/data_model/anch_tot.rds", sep = ""))
data_model <- data_model %>% filter(season == "spring")



ggplot(study_domain_sf) + geom_sf() +
  geom_point(data = data_model %>% filter(!is.na(indice)),
             aes(x = Xgd, y = Ygd, color = indice), size = 1) +
  labs(x = "", y = "", fill = "Occurence", 
       title = "Carte de présence de l'anchois dans le Golfe de Gascogne par Pelgas") +
  facet_wrap( ~ year, ncol = 7, nrow = 3) +
  theme_bw()

ggplot(study_domain_sf) + geom_sf() +
  geom_point(data = data_model %>% filter(!is.na(biomass)), 
  aes(x = Xgd, y = Ygd, color = biomass), size = 1) +
  labs(x = "", y = "", fill = "Biomasse en kg", 
       title = "Carte de densité de l'anchois dans le Golfe de Gascogne par Pelgas") +
  facet_wrap( ~ year, ncol = 7, nrow = 3) +
  theme_bw()

ggplot(study_domain_sf) + geom_sf() +
  geom_tile(data = data_model %>% filter(is.na(indice)),
            aes(x = Xgd, y = Ygd, color = presence), size = 1) +
  labs(x = "", y = "", fill = "Occurence", 
       title = "Carte de présence de l'anchois dans le Golfe de Gascogne par la Pêche") +
  facet_wrap( ~ year, ncol = 7, nrow = 3) +
  theme_bw()
```

# Mesh construction
```{r}

plot(mesh_GdG)

ggplot(study_domain_sf) + 
  geom_sf() + 
  theme_bw()


```

For the model we choose the mesh_GdG to allow comparaison

# Create the space weigth and add it
```{r}
# It is necessary to represent the poisson point process

space_weight <- create_space_weight(mesh_GdG, study_domain_sf)

### this step is used only if you study_domain is not a single polygon
### so some cells have been cut in several parts and we sum together this part
ID <- c()
area <- c()
for (I in 1:mesh_GdG$n){
  val <- space_weight %>%
    filter(ID == I) %>%
    colSums(.)
  
  ID <- c(ID,I)
  area <- c(area, val[2])
}
space_weight <- data.frame(ID,area)

df_weight <- add_weight(mesh_GdG, space_weight, data_model)

data_model <- data_model %>%
  mutate(weight = ifelse(survey == "PECHE", 0, 1)) %>%
  st_cast()

data_model <- rbind(data_model,df_weight)

saveRDS(data_model, paste(local_path,"code/data_model/anch_tot.rds", sep = ""))


```


# Building the SPDE model on the mesh

Now we use the inla.spde2.pcmatern() function to build the SPDE model
and specify Penalised Complexity priors for the parameters of the Matérn
field.

```{r}

spde <- inla.spde2.pcmatern(
  mesh = mesh_GdG, 
  alpha = 1.5,
  prior.range = c(0.6, 0.1), # P(range < 0.6) = 0.1 (0.6 = 1.8/3)
  prior.sigma = c(log(5) / 3, 0.01) # P(sigma > log(5) / 3) = 0.01
  ) 

```

# Index set

Now we construct the index set for the latent spatio-temporal Gaussian
model

```{r}
timesn <- max(data_model$year) - min(data_model$year) + 1 #attention max(timesn) doit être = max(group)
sn <- 1 # there is one season

field.z.idx <- inla.spde.make.index(name = 'x', 
  n.spde = spde$n.spde, n.group = 1)

field.zc.idx <- inla.spde.make.index(name = 'xc', 
  n.spde = spde$n.spde, n.group = timesn)

field.y.idx <- inla.spde.make.index(name = 'u', 
  n.spde = spde$n.spde, n.group = 1)

field.p.idx <- inla.spde.make.index(name = 'p', 
  n.spde = spde$n.spde, n.group = 1)

lengths(field.z.idx)
lengths(field.zc.idx)
lengths(field.y.idx)
lengths(field.p.idx)
```
# Projection matrix

Now we build a projection matrix A that projects the spatio-temporal
continuous Gaussian random field from the observations to the mesh
nodes.

```{r}
coo <- data_model %>%
  st_drop_geometry() %>%
  dplyr::select(Xgd, Ygd)

group <- data_model$year - min(data_model$year) + 1
A <- inla.spde.make.A(mesh = mesh_GdG, 
                      # must be a matrix
                      loc = coo %>%
                        as.matrix(), 
                      group = group
                      )

```


# Prepare the prediction for Gaussian field
```{r}
grille_pred <- sf::st_read("C:/Users/ahebertb/Desktop/StageAurel/20230724_envdata_bob.gpkg", layer = "2021")

grille_pred <- grille_pred %>% st_transform(., 4326) %>% 
  filter(season == "summer")
  
ggplot(grille_pred)+geom_sf()

coord_pred <- grille_pred %>% st_centroid(.) %>% st_coordinates(.)

A_pred <- inla.spde.make.A(mesh = mesh_GdG, 
                      # must be a matrix
                      loc = coord_pred %>%
                        as.matrix(), 
                      group = sn*timesn,
                      )
  
```



# Stack with data 

```{r}
# standardize covariates
std <- function(x, xref = NULL) {
  if(is.null(xref)) {
    out <- (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  } else {
    out <- (x - mean(xref, na.rm = TRUE)) / sd(xref, na.rm = TRUE)
  }
  return(out)
}

#stack
stk.z  <- inla.stack(
  data = list(Y = cbind(as.vector(data_model$indice), NA, NA), e = data_model$weight , link = 1), 
  A = list(A, 1),
  effects = list(c(field.z.idx, field.zc.idx), 
                 z.intercept = rep(1, nrow(data_model))
  ),
  tag = 'occ'
) 

stk.y <- inla.stack(
  data = list(Y = cbind(NA, as.vector(data_model$biomass),NA), e = data_model$weight, link = 2),
  A = list(A, 1),
  effects = list(c(field.y.idx, field.zc.idx), 
                 y.intercept = rep(1, nrow(data_model))
  ), 
  tag = 'dens'
) 

stk.p <- inla.stack(
  data = list(Y = cbind(NA, NA, as.vector(data_model$presence)), e = data_model$weight, link = 3),
  A = list(A, 1),
  effects = list(c(field.p.idx, field.zc.idx), 
                 p.intercept = rep(1, nrow(data_model))
  ), 
  tag = 'pres'
)

#stack prediction
empty.pred <- rep(NA, length(coord_pred)/2)

stk.z.pred <- inla.stack(
  data = list(Y = cbind(empty.pred, NA, NA),
              e = rep(1, length(coord_pred)/2), link = 1), 
  A = list(A_pred, 1),
  effects = list(c(field.z.idx, field.zc.idx), 
                 z.intercept = rep(1, nrow(A_pred))
  ),
  tag = 'occ.pred'
) 

stk.y.pred <- inla.stack(
  data = list(Y = cbind(NA, empty.pred, NA),
              e = rep(1, length(coord_pred)/2), link = 2),
  A = list(A_pred, 1),
  effects = list(c(field.y.idx, field.zc.idx), 
                 y.intercept = rep(1, nrow(A_pred))
  ), 
  tag = 'dens.pred'
) 

stk.p.pred <- inla.stack(
  data = list(Y = cbind(NA, NA, empty.pred), e = rep(0, length(coord_pred)/2), link = 3),
  A = list(A_pred, 1),
  effects = list(c(field.p.idx, field.zc.idx), 
                 p.intercept = rep(1, nrow(A_pred))
  ), 
  tag = 'pres.pred'
)

stk.all <- inla.stack(stk.z, stk.y, stk.p,
                      stk.z.pred, stk.y.pred, stk.p.pred)
```

# temporal autoregressive parameter
```{r}
h.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
```
control.group = list(model = "ar1", hyper = rprior) specifes that across time, the process evolves according to an AR(1) process where the prior for the autocorrelation parameter a is given by rprior. We define rprior with the prior "pccor1" which is a PC prior for the autocorrelation parameter a where a=1 is the base model. Here we assume P(a>0)=0.9.

# Prior
```{r}
#intercept prior
prior.fixed <- list(mean.intercept = 0.0, prec.intercept = 1 / (log(10) / 2)^2, mean = 0.0, prec = 1 / (log(5) / 2)^2) 

cinla <- list(strategy = 'simplified.laplace', int.strategy = 'eb',  control.vb=list(emergency= 30))

bprior <- list(prior = 'gaussian', param = c(0, 1))
```
For the shared parameter β we can set a prior based on some knowledge about the correlation between the rain occurrence and the rainfall amount. We assume a N(0, 1) prior for the bprior.

#Model formula

```{r}
formula.joint <- Y ~ -1 + 
    z.intercept + y.intercept + p.intercept +
    f(x, model = spde, group = x.group, control.group = list(model = "iid")) + #binom field
    f(xc, model = spde, fixed = FALSE, #hyper = list(beta = bprior), 
      group = xc.group,
      control.group = list(model = 'ar1', hyper = h.spec)) + #shared field
    f(u, model = spde, group = u.group, control.group = list(model = "iid")) + #gamma field
    f(p, model = spde, group = p.group, control.group = list(model = "iid")) #PPP field

```
We can define the model formula for the model specified. We use the spde object to define the model of the space-time component together with the definition of the prior for the AR(1) temporal dynamics. In order to define the β parameter of the shared space-time component we set fixed = FALSE to estimate β and insert its prior.


# inla() call
```{r}
res.jo <- inla(formula.joint, family = c("binomial", "lognormal", "poisson"), 
  data = inla.stack.data(stk.all), 
  control.predictor = list(A = inla.stack.A(stk.all), link = link, compute = TRUE),
  control.compute = list(dic = FALSE, waic = TRUE, cpo = TRUE,
    config = TRUE, residuals = TRUE), control.inla = cinla, control.fixed = prior.fixed, 
  verbose = TRUE, E = inla.stack.data(stk.all)$e)

```

# Show the result
```{r}
summary(res.jo)

saveRDS(res.jo, file = paste(local_path,"/code/model/result_anch_spring_894",".rds", sep = ""))
res.jo <- readRDS(paste(local_path,"/code/model/result_anch_spring_894",".rds", sep = ""))

res.jo$mode$mode.status 
```
