---
title: "Model_Pelgas"
output: html_document
date: "2023-04-25"
editor_options: 
  markdown: 
    wrap: 72
---

All the syntax starting with a \# in a chunk do not run

We prepare the data we will use in this code

```{r}
grid_projection <- "+proj=longlat +datum=WGS84" #grid_projection is a valid CRS specification in the PROJ.4 format

load("C:/Users/ahebertb/Desktop/StageAurel/code_florian/mesh_GdG.Rdata")
load("C:/Users/ahebertb/Desktop/StageAurel/code_florian/study_domain_sp.RData")


```

#To use the crs() function, you first need to convert your
study_domain_sp object to an sf object. \# Convert study_domain_sp to an
sf object

```{r}
study_domain_sf <- st_as_sf(study_domain_sp)
st_crs(study_domain_sf) <- 4326
world <- ne_download(scale = 10, type = 'countries', returnclass = "sf")
ggplot(study_domain_sf)+geom_sf()+theme_bw()
```

# Importation of the data

```{r}
path_data <- "C:/Users/ahebertb/Desktop/StageAurel/code/data/tidy/"
sard_PEL_2018 <- readRDS(paste(path_data,"sard_2018_PEL.RDS",sep = ""))
SST_2018 <- readRDS(paste(path_data,"SST_2018.RDS",sep = ""))
```

# First transformation of the Pelgas dataset

```{r}
df.sard_PEL_2018 <- st_as_sf(sard_PEL_2018, coords = c("LONG","LAT"))
st_crs(df.sard_PEL_2018) <- 4326

df.SST_2018 <- st_as_sf(SST_2018, coords = c("long","lat"))
st_crs(df.SST_2018) <- 4326
```

# We want to keep the data of the domain of study

```{r}
location <- st_intersects(study_domain_sf,df.sard_PEL_2018)
obs <- sard_PEL_2018[location[[1]],]
for (polygon in (2:length(location))){
  obs <- rbind(obs,sard_PEL_2018[location[[polygon]],])
}


ggplot(study_domain_sf)+geom_sf()+
  geom_point(data = obs, aes(x = LONG, y = LAT), size = 0.5)+theme_bw()



location <- st_intersects(study_domain_sf,df.SST_2018)
sst <- SST_2018[location[[1]],]
for (polygon in (2:length(location))){
  sst <- rbind(sst,SST_2018[location[[polygon]],])
}

sst <- (cbind(sst$long, sst$lat, sst[,7], rep(2018,143)))
sst <- as.data.frame(sst)
names(sst) <- c("Xgd","Ygd","SST","Year")

sst$Xgd <- as.numeric(sst$Xgd)
sst$Ygd <- as.numeric(sst$Ygd)
sst$SST <- as.numeric(sst$SST)
sst$Year <- as.numeric(sst$Year)


ggplot(study_domain_sf)+geom_sf()+
  geom_point(data = sst, aes(x = Xgd, y = Ygd), size = 0.5)+theme_bw()

```

# We want associate all the data of the domain of study
```{r}
path_map <- paste(path.output,"fusion/sard/",sep = "")
source(paste(path_function,"prepare_grid_pel_jun.R",sep = ""))

Grid2 <- mosaic_gridDataframe(obs,"year", 
                               path_map,"indice",
                                "indice")
Grid2 <- Grid2 %>% mutate( presence = Zvalue)
#the grid is used to concatenate the value of Pelgas 

df <- merge(sst,Grid2,by=c("Xgd","Ygd","Year"))

```



# Mesh construction

We compare a random mesh contruction with the mesh_GdG of Florian's work

```{r}
coo <- cbind(df$Xgd,df$Ygd)

#it's a try of my own mesh to check how it work
bnd <- inla.nonconvex.hull(st_coordinates(study_domain_sf))
mesh <- inla.mesh.2d(loc = coo, boundary = bnd,
  cutoff=0.11,
  max.edge=c(0.1,0.45),min.angle=c(21,21),crs = grid_projection
)

plot(mesh)

plot(mesh_GdG)
points(coo, col = "red")

plot(mesh_GdG)
points(cbind(sst$Xgd,sst$Ygd), col = "red")
```

For the model we choose the mesh_GdG to allow comparaison

# Building the SPDE model on the mesh

Now we use the inla.spde2.pcmatern() function to build the SPDE model
and specify Penalised Complexity priors for the parameters of the Matérn
field.

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh_GdG,
  prior.range=c(0.01,0.1),
  prior.sigma=c(100,0.1)
)
#les paramètres sont ceux de Florian
```

# Index set

Now we construct the index set for the latent spatio-temporal Gaussian
model

```{r}
timesn <- length(unique(df$Year))
indexs <- inla.spde.make.index(
  "s", n.spde = spde$n.spde, n.group = timesn
)
lengths(indexs)
```

# Projection matrix

Now we build a projection matrix A that projects the spatio-temporal
continuous Gaussian random field from the observations to the mesh
nodes.

```{r}
group <- df$Year - min(df$Year) + 1
A <- inla.spde.make.A(mesh = mesh_GdG, loc = coo, group = group)
```


# Stack with data 

```{r}
stk.pres <- inla.stack(
  tag = "occdata",
  data = list(y = df$presence),
  A = list(1,A),
  effects = list( t=df$SST, s=indexs)
)


```

#Model formula

```{r}
rprior <- list(theta =list(prior = "pc.cor1", param = c(0, 0.9)))

formula <- y ~ 0 + t + f(s, model = spde, group = group,
                         control.group = list(model = "ar1", hyper = rprior))
```

# inla() call
```{r}

res <- inla(formula, family = c("binomial"), 
  data = inla.stack.data(stk.pres), 
  control.predictor = list(compute = TRUE, A = inla.stack.A(stk.pres)),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
  )
```


